---
title: "Impact of weather events using storm database"
author: "KHAWLA"
date: "2023-04-20"
output:
  html_document:
    df_print: paged
---
Reproducible Research Project 2
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in **fatalities**, **injuries**, and **property damage**, and preventing such outcomes to the extent possible is a key concern.
This study explores the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.
Tornadoes are wreaking havoc among the catastrophes with the most fatalities and injuries while property damage is due to floods.

# Data

The events in the database start in the year 1950 and end in November 2011.

* **DataSet**: [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
* **Documentation**: [National Weather Service](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

# Data Processing

### Loading storm data
```{r pressure, echo=FALSE}
setwd("C:/Users/KHAWLA/Downloads/")
```
```{r, echo=TRUE}
file.url <- ("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2")
download.file(file.url, destfile = "repdata_data_StormData.csv.bz2")
data.storm <- read.csv(bzfile("repdata_data_StormData.csv.bz2")) 
```

### Data Preview

```{r}
str(data.storm)
summary(data.storm)
head(data.storm)
names(data.storm) <- tolower(names(data.storm))
```
The data set contains 37 variables and 902297 observations.

### Data Preparation

```{r}
library(dplyr)
storm <- data.storm %>%
  select(c("evtype","fatalities","injuries","propdmg","propdmgexp","cropdmg","cropdmgexp"))
head(storm)
summary(storm)
str(storm)
View(storm)
sum(is.na(storm))
```
We'll use only 7 variables:

**evtype**: type of event

**fatalities**: number of fatalities for each event

**injuries**: number of injuries for each event

**propdmg**: property damage

**propdmgexp**: crop damage

**cropdmg**: property damage exponents (in B, M, K and H)

**cropdmgexp**: crop damage exponents (in B, M, K and H)

```{r}
rm(data.storm)
```
we remove the raw data set and keep on working on the *storm* data set

### Data Cleaning

```{r}
sum(unique(storm$fatalities))
sum(unique(storm$injuries))
sum(unique(storm$propdmg))
unique(storm$propdmgexp)
unique(storm$cropdmgexp)
```
we look for any unusual data 
```{r}
storm$propdmgexp[storm$propdmgexp %in% c("","+","-","?")] <- "0"
storm$propdmgexp[storm$propdmgexp %in% c("B")] <- "9"
storm$propdmgexp[storm$propdmgexp %in% c("M","m")] <- "6"
storm$propdmgexp[storm$propdmgexp %in% c("K")] <- "3"
storm$propdmgexp[storm$propdmgexp %in% c("H","h")] <- "2"
```

```{r}
storm$cropdmgexp[storm$cropdmgexp %in% c("","?")] <- "0"
storm$cropdmgexp[storm$cropdmgexp %in% c("B")] <- "9"
storm$cropdmgexp[storm$cropdmgexp %in% c("M","m")] <- "6"
storm$cropdmgexp[storm$cropdmgexp %in% c("K","k")] <- "3"
```

```{r}
storm$propdmgtotal <- storm$propdmg*(10^as.numeric(storm$propdmgexp))
storm$cropdmgtotal <- storm$cropdmg*(10^as.numeric(storm$cropdmgexp))
storm$dmgtotal <- storm$propdmgtotal+ storm$cropdmgtotal
```

we found that the last two variables do not have clean data so we clean it.

```{r}
storm_by_event <- storm %>%
  group_by(evtype) %>%
  summarize(sum_fatalities = sum(fatalities),
            sum_injuries = sum(injuries),
            sum_propdmg = sum(propdmgtotal),
            sum_cropdmg = sum(cropdmgtotal),
            sum_totaldmg = sum(dmgtotal))
head(storm_by_event)
```
we summarized the data by events showing each event with its summarized fatalities, injuries and damages.

1. Across the United States, which types of events (as indicated in the **EVTYPE** variable) are most harmful with respect to population health?
```{r}
# Events that caused most fatalities
storm_fatalities <- arrange(storm_by_event, desc(sum_fatalities))
fatalitydata <- head(storm_fatalities)
fatalitydata

# Events that caused most injuries
storm_injuries <- arrange(storm_by_event, desc(sum_injuries))
injurydata <- head(storm_injuries)
injurydata
```
we retain only the first 6 events that caused the most fatalities and injuries.

2. Across the United States, which types of events have the greatest economic consequences?
```{r}
# Events with greatest economic consequences
storm_damage <- arrange(storm_by_event, desc(sum_totaldmg))
damagedata <- head(storm_damage)
damagedata
```
we retain only the first 6 events that caused the most economic consequences.

# Results
```{r}
library(ggplot2)
```
**FATALITIES**
```{r}
ggplot(fatalitydata, aes(reorder(evtype, -sum_fatalities),sum_fatalities))+
  geom_bar(stat = "identity",color= "black", fill="purple")+
  geom_text(aes(label = sum_fatalities), nudge_y = 200)+
  ggtitle("Events that caused most fatalities")+
  xlab("Event type") + ylab("Total of fatalities") +
  theme_classic()
```

it looks like **Tornadoes** causes the most fatalities with 6533 and exceeds by far **excessive heat** which registered 1903 fatalities.

**INJURIES**
```{r}
ggplot(injurydata, aes(reorder(evtype, -sum_injuries),sum_injuries))+
  geom_bar(stat = "identity",color= "black", fill="pink")+
  geom_text(aes(label = sum_injuries), nudge_y = 4000)+
  ggtitle("Events that caused most injuries")+
  xlab("Event type") + ylab("Total of injuries") +
  theme_classic()
```

it looks like **Tornadoes** causes the most injuries with 91346 and exceeds by far **TSTM Wind** which registered 6957 injuries.

**ECONOMIC CONSEQUENCES/DAMAGES**
```{r}
library(tidyverse)
```
```{r}
head(damagedata)
damage <- damagedata %>%
  gather(key = "Type", value = "TotalDamage",c("sum_propdmg","sum_cropdmg")) %>%
  select(evtype, Type, TotalDamage)

damage
damage$Type[damage$Type %in% c("sum_propdmg")] <- "property damage"
damage$Type[damage$Type %in% c("sum_cropdmg")] <- "crop damage"
damage
damage$evtype <- with(data = damage, reorder(evtype, -TotalDamage))

ggplot(damage, aes(x = evtype, y = TotalDamage, fill = Type))+
  geom_bar(position = "stack" , stat = "identity")+
  ggtitle("Events that caused most damages")+
  xlab("Event type") + ylab("Total of damages") +
  theme_classic()+ theme(axis.text.x = element_text(angle = 10))
```

it looks like **Floods** cause the most damage in terms of crop and property damages followed by **Hurricanes** and **Tornadoes**
